\chapter{Particle Physics}
\restartlist{enumerate}
\section{Accelerators}
\begin{enumerate}
    \item The available energy in a particle collision is defined as the available kinetic energy unless otherwise specified.
    \item The threshold energy is the minimum energy needed by one particle to produce another.
    \item Generally, energies required in a center-of-momentum frame are much lower than those required in lab frames.
\end{enumerate}
\section{Standard Model}
\begin{enumerate}
    \item The standard model consists of:
    \begin{itemize}
        \item 6 quarks: up, down, charm, strange, top/truth, bottom/beauty
        \item 6 leptons: electron, muon, tau, and their corresponding neutrinos
        \item 6 gauge bosons: gluon, W\textsuperscript{+}, W\textsuperscript{-}, Z, photon, graviton
    \end{itemize}
    \item A hadron is a composite particle made of quarks. There are only 3 ways to combine quarks: baryons (3 quarks), antibaryons (3 antiquarks) and mesons (quark-antiquark pair).
    \item Bosons are particles with integer spins; fermions are particles with half-integer spins.
    \item All fermions obey Pauli's exclusion principle.
    \item All quarks and leptons half spin half, while gauge bosons have spin 1. Protons and neutrons both have spin half as well.
    \item Up, charm and truth have electronic charge \(+2/3e\) while down, strange and beauty have electronic charge \(-1/3e\).
\end{enumerate}
\section{Conservation Laws}
\begin{enumerate}
    \item Baryon number is simply the number of baryons, where antibaryons are considered negative baryons. In all interactions, baryon number is conserved.
    \item Each of the 6 quarks have their own number i.e. up, down, charm, strangeness, topness, bottomness. All are conserved in all interactions except weak interactions.
    
    Note that up, charm and truth are assigned \(+1\) in their own numbers while down, strange and beauty are assigned \(-1\). Antiquarks are given the opposite.
    \item Each of the lepton generations have their own conservation numbers, where leptons are given \(+1\) while antileptons are given \(-1\). This is conserved in all interactions.
    
    Neutrinos and their corresponding particles are conserved in the same lepton number.
\end{enumerate}